\section{Introduction}

The rapid proliferation of text-to-image (T2I) diffusion models~\citep{rombach2022high,ramesh2022hierarchical} has enabled users to generate photorealistic images from natural language descriptions. While this capability has transformative potential for creative industries, it simultaneously raises critical safety concerns: these models can be exploited to generate harmful content including sexual nudity, violence, hate imagery, and other inappropriate material~\citep{rando2022red,schramowski2023safe,qu2023unsafe}. The challenge of \emph{concept erasure}---suppressing specific unsafe concepts while preserving general generative capability---has thus emerged as a pressing research problem.

Existing safety mechanisms can be broadly categorized into two paradigms. \textbf{Training-based approaches}~\citep{gandikota2023erasing,kim2023towards,gandikota2024unified,gong2024reliable} permanently alter model parameters to remove harmful concepts. While effective at targeted suppression, these methods suffer from several limitations: (1) each concept requires separate fine-tuning, incurring substantial computational cost; (2) parameter modification risks catastrophic forgetting of benign concepts; and (3) once modified, the model remains constrained even when generating entirely safe content.
\textbf{Training-free inference-time guidance}~\citep{schramowski2023safe,brack2023sega,yoon2024safree} preserves model parameters and intervenes dynamically during generation. However, existing methods apply suppression globally across all spatial locations and timesteps, leading to over-correction of benign regions, semantic drift, and degraded image quality---even when unsafe content is spatially localized or entirely absent.

These limitations stem from a shared fundamental issue: \emph{existing methods lack the ability to determine when, where, and how to intervene in a context-dependent manner}. Unsafe content is often sparse, implicit, or localized within specific image regions, yet current approaches apply uniform intervention regardless of the actual generation context.

To address this, we propose \textbf{Example-Based Spatial Guidance (\ours{})}, a training-free concept erasure framework. Our core design choice is that \emph{all concept-specific knowledge lives in a small exemplar pack}---a handful of paired target/anchor prompts and images organized by sub-category---which the user is free to author, extend, or replace. From this pack alone, \ours{} decides \textbf{when} to intervene (a Concept Alignment Score, CAS, gates by cosine similarity between the denoising direction and the exemplar-defined target direction, \S\ref{subsec:when}) and \textbf{where} to intervene (a dual-probe mask combines cross-attention with CLIP exemplar similarity, \S\ref{subsec:where}); the suppression step (\S\ref{subsec:how}) is a plain mask-gated adjustment to the CFG prediction, offered in \emph{two complementary modes}---anchor inpainting and hybrid guidance---that span a compute$\times$anchor-cleanliness trade-off and are reported as two first-class rows throughout.

\paragraph{Why this matters: training-free and user-customizable.}
Because only the exemplar pack carries concept semantics, \ours{} inherits two practical strengths that differentiate it from both fine-tuning methods and prior inference-time baselines:
\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
    \item \textbf{Training-free} (\S\ref{subsec:training_free}). No weights are updated and no auxiliary classifier is trained, eliminating the catastrophic-forgetting and per-concept-retraining costs incurred by ESD/SDD/RECE.
    \item \textbf{User-customizable, few-shot adaptable.} A concept is defined by $\sim\!16$ prompts and $\sim\!16$ images. If the target distribution shifts---for example, to the MMA benchmark's metaphorical attacks~\citep{zhang2024metaphor}---a few in-distribution exemplars can be dropped into the pack and embeddings recomputed in seconds, with no gradient step. The safety layer is thus \emph{editable}, not a fixed artifact.
\end{itemize}

Importantly, we discover that our dual-probe mechanism exhibits \emph{concept-specific dominance}: the text probe excels at detecting explicitly named concepts (e.g., nudity, where adversarial prompts still contain identifiable keywords), while the image probe dominates for visually implicit concepts (e.g., violence and illegal activity, where metaphorical or indirect language evades text-based detection). Their union consistently achieves the best performance across all categories, validating the complementary design.

\paragraph{Families as the default.} Unsafe concepts are rarely monolithic: \emph{nudity} spans naked body, sexual act, exposed parts, and suggestive pose; \emph{violence} spans blood/gore, weapon attack, brutal action, and dismemberment. Pooling all exemplars into one averaged embedding collapses this diversity. \ours{} therefore uses \emph{family-grouped exemplars} ($F{=}4$ families, $K{=}4$ paired target/anchor exemplars per family) as the first-class concept representation throughout CAS, the probes, and the guidance terms. Family grouping (1) improves coverage on diverse concepts (MJA-Violent +21\%p, MJA-Illegal +20\%p vs.\ the single-pooled variant), (2) sharpens spatial masks, and (3) reduces destructive interference in simultaneous multi-concept erasure by routing each concept's suppression to $F$ independent sub-directions rather than a single collapsed vector.

We evaluate \ours{} extensively across seven safety categories from the I2P benchmark~\citep{schramowski2023safe} (nudity, violence, harassment, hate, shocking, illegal activity, and self-harm), multiple adversarial prompt datasets (Ring-A-Bell~\citep{tsai2024ring}, MMA-Diffusion~\citep{yang2024mma}, P4D-N~\citep{chin2024prompting4debugging}, and UnlearnDiff~\citep{zhang2024generate}), and the MJA adversarial benchmark~\citep{zhang2024metaphor} which employs metaphor-based jailbreaking attacks across four categories (sexual, violent, disturbing, illegal; 100 prompts each). Our contributions are:

\begin{enumerate}[leftmargin=1.5em,itemsep=2pt]
    \item We propose \ours{}, a \emph{training-free} concept-erasure framework whose only concept-specific object is a small, user-authored exemplar pack. No weights are updated, no classifier is trained, and no dataset is collected.
    \item We show that this exemplar pack enables \emph{few-shot adaptation} to shifted adversarial distributions (\eg, MMA-style metaphorical attacks): adding a handful of in-distribution exemplars and recomputing embeddings suffices---no gradient step, no retraining.
    \item We derive from the pack two lightweight but effective signals---a Concept Alignment Score (When) and a dual text/image probe (Where)---and expose their \emph{concept-specific dominance}: text probes for explicitly-named concepts (nudity), image probes for visually-implicit ones (violence, illegal).
    \item We use \emph{family-grouped exemplars} as the default pack representation, improving coverage on semantically diverse concepts (MJA Violent $+21$\%p, MJA Illegal $+20$\%p) and reducing destructive interference in simultaneous multi-concept erasure.
    \item We present the \emph{How} stage as two complementary modes---\emph{anchor inpainting} (subtract-only, compute-light) and \emph{hybrid guidance} (subtract-plus-anchor-attract, stronger correction)---that are reported together as first-class rows in every results table. They span a compute$\times$anchor-cleanliness trade-off rather than competing for a single slot.
    \item We demonstrate state-of-the-art training-free results across seven I2P categories, four nudity benchmarks, and the MJA adversarial benchmark---outperforming SAFREE on 10/14 benchmarks with gains up to $+49.7$\%p---while preserving image quality (FID 6.78 on COCO, VQA drop $<0.05$).
    \item We validate our evaluation protocol via MJ-Bench~\citep{chen2024mjbench}, establishing Qwen3-VL~\citep{bai2025qwen3vl} as a reliable automated safety judge ($92.3\%$ cross-validator agreement).
\end{enumerate}
