\section{Experiments}
\label{sec:exp}

\subsection{Experimental Setup}

\paragraph{Base Model.} We use Stable Diffusion v1.4~\citep{rombach2022high} (\texttt{CompVis/stable-diffusion-v1-4}) as the base T2I model, with the safety checker disabled to ensure a fair evaluation of concept erasure methods. All images are generated at 512$\times$512 resolution with 50 DDIM steps, $s=7.5$ guidance scale, and a fixed seed of 42. Cross-backbone validation (\S\ref{subsec:cross_backbone}) additionally uses SD~3 (MMDiT, velocity parameterization) and FLUX.1-dev (DiT, flow matching with embedded guidance).

\paragraph{Datasets.} For nudity erasure, we evaluate on three adversarial prompt datasets: \textbf{Ring-A-Bell}~\citep{tsai2024ring} (79 prompts), \textbf{UnlearnDiff}~\citep{zhang2024generate} (142 prompts), and \textbf{MMA-Diffusion}~\citep{yang2024mma} (1,000 prompts). Additional results on P4D-N~\citep{chin2024prompting4debugging} (151 prompts) are reported in the appendix. For multi-concept evaluation, we use I2P~\citep{schramowski2023safe} subsets for each safety category: violence (756), harassment (824), hate (231), shocking (856), illegal activity (727), and self-harm (801). We additionally evaluate on the \textbf{MJA} (Metaphor-based Jailbreaking Attack)~\citep{zhang2024metaphor} benchmark, which employs indirect metaphorical adversarial prompts across four categories (sexual, violent, disturbing, illegal; 100 prompts each), providing a more challenging evaluation than standard I2P prompts. For image quality assessment, we generate images from 250 COCO~\citep{lin2014microsoft} captions ($\times$4 samples = 1,000 images) and compute FID against baseline SD v1.4 outputs. We further report VQAScore~\citep{lin2024evaluating} as a prompt faithfulness metric.

\paragraph{Baselines.} We compare against: (1)~\emph{Fine-tuning methods}: ESD~\citep{gandikota2023erasing}, SDD~\citep{kim2023towards}, and RECE~\citep{gong2024reliable}; (2)~\emph{Training-free methods}: SLD-Max~\citep{schramowski2023safe}, SAFREE~\citep{yoon2024safree}, and our prior ASCG~\citep{gil2026ascg}. We use SAFREE as our primary comparison for multi-concept evaluation, as it is the strongest existing training-free baseline with native multi-concept support.

\paragraph{Evaluation Metrics.} We use Qwen3-VL-8B~\citep{bai2025qwen3vl} as the primary safety evaluator, classifying each generated image into four categories: \textbf{Safe}, \textbf{Partial} (minor exposure), \textbf{Full} (explicit unsafe content), and \textbf{NotRelevant} (no human/relevant subject). We report:
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
    \item \textbf{SR} (Safe Rate): $(\text{Safe} + \text{Partial}) / \text{Total}$ --- higher is better
    \item \textbf{Full}: fraction of fully unsafe images --- lower is better
    \item \textbf{FID}~\citep{heusel2017gans}: Fr\'{e}chet Inception Distance on COCO --- lower is better
    \item \textbf{CLIP Score}~\citep{hessel2021clipscore}: text-image alignment on COCO --- higher is better
\end{itemize}
We validate our VLM evaluator through MJ-Bench~\citep{chen2024mjbench} benchmarking (\S\ref{subsec:evaluator}).

% ===========================================================================
\subsection{Main Results: Nudity Erasure}
\label{subsec:main_results}

Table~\ref{table:main} presents the comprehensive comparison on nudity erasure across four adversarial prompt datasets and COCO image quality metrics.

\input{table/main}

\paragraph{Comparison with Training-free Methods.}
Among training-free approaches, \ours{} achieves 87.3\% SR on Ring-A-Bell, substantially outperforming pure SAFREE (50.6\%), SLD-Max (68.4\%), and even the strengthened baselines SAFREE+Safe\_Denoiser (72.2\%) and SAFREE+SGF (65.8\%). On P4DN, the gap is even larger: \ours{} achieves 94.7\% vs.\ SAFREE+SGF's 59.9\%. On the challenging MJA-Sexual benchmark with metaphorical adversarial prompts, \ours{} reaches 84.0\% compared to SAFREE's 48.0\% (+36.0\%p). Notably, Safe\_Denoiser and SGF are evaluated using SAFREE as their base following their official codebases, yet \ours{} as a standalone method still outperforms these compound baselines.

\paragraph{Comparison with Fine-tuning Methods.}
Fine-tuning methods (ESD, SDD) achieve strong nudity erasure but at significant cost: high FID scores indicate quality degradation on benign prompts, and they require retraining for each new concept. \ours{} achieves competitive SR (87.3\% vs.\ SDD's 88.6\% on RAB) while maintaining FID 6.78, and generalizes to new concepts without any training.

\paragraph{Image Quality.}
\ours{} achieves FID 6.78, substantially lower than SLD-Max (18.46) and SAFREE (8.96), confirming that CAS gating effectively prevents quality degradation on benign prompts.

% ===========================================================================
\subsection{Multi-Concept Erasure}
\label{subsec:multi_results}

Beyond nudity, \ours{} generalizes to other unsafe concepts without any concept-specific training---only the exemplar prompts and target keywords are changed. We evaluate on the MJA benchmark, which provides more challenging adversarial prompts than standard I2P datasets. Table~\ref{table:multi} presents per-concept erasure results.

\input{table/multi_concept}

\paragraph{Per-Concept Results.}
Reporting both How modes per concept (\S\ref{subsec:how}) reveals a complementary pattern: hybrid guidance wins on concepts with a visually crisp paired anchor---MJA-Sexual (86.0\%, +38.0\%p vs.\ SAFREE) and MJA-Disturbing (99.0\%, +4.0\%p, vs SAFREE 95.0\%)---while anchor inpainting wins on MJA-Violent (56.0\%, +4.0\%p), where ``peaceful scene'' as the paired anchor is ambiguous and hybrid's chase-the-anchor objective becomes unstable (hybrid 29.0\%). On MJA-Illegal, SAFREE (77.0\%) still holds an edge over both modes (anchor 56.0\%, hybrid 72.0\%); the image-dominant nature of the concept (Appendix~\ref{app:probe_dominance}) combined with ambiguous anchors (\textit{drug use} $\to$ \textit{legal activity} does not correspond to a clean visual pair) is the shared challenge across all training-free methods on this category. The key point is that \emph{which} mode wins is concept-dependent; the paper therefore reports both and leaves the choice to the deployer.

\paragraph{Simultaneous Multi-Concept Erasure.}
Table~\ref{table:dual} reports dual-concept erasure on I2P-Sexual (931 prompts) using the SAFREE-style per-concept breakdown: for each method, we show the Qwen3-VL class distribution \emph{separately} for each of the two jointly-erased concepts (left: Nudity, right: 2nd concept) alongside SR~$=$~(Safe~$+$~Partial)$/$Total. With family-grouped packs as the default, Nudity~$+$~Shocking retains $83.7\%$ nudity SR while also reaching $97.5\%$ on Shocking---close to the single-concept references ($93.1\%$ and $97.3\%$ respectively), i.e.\ little destructive interference. Because each concept is decomposed into $F$ semantically independent anchors, simultaneous suppression is routed per sub-category instead of through a single collapsed direction, which makes family grouping particularly attractive for \emph{multi-concept} deployment, not just per-concept gains. Results for the remaining three pairs (Nudity~$+$~Violence, Nudity~$+$~Harassment, Nudity~$+$~Hate) are reported in Appendix~\ref{app:dual_pairs} (Table~\ref{table:dual_all}) in the same format.

\input{table/dual_concept}

% ===========================================================================
\subsection{Ablation Study}
\label{subsec:ablation}

We ablate the three core components of \ours{} on Ring-A-Bell and MMA-Diffusion.

\input{table/ablation}

\paragraph{Probe Mode Comparison.}
Table~\ref{table:probe} compares text-only, image-only, and dual (both) probes across nudity datasets and MJA-Sexual, run under anchor inpainting for clean attribution (the probes and the How mode are orthogonal components of \ours{}). Text-only achieves 86.1\% on RAB and 94.0\% on P4DN, while image-only reaches 84.8\% and 90.7\% respectively. The dual probe consistently achieves the best or tied-best results (87.3\% RAB, 94.7\% P4DN). On MJA-Sexual with adversarial prompts, image probe (86.0\%) slightly outperforms text probe (85.0\%), because metaphorical prompts avoid explicit keywords. This \emph{concept-specific probe dominance}---text for explicit concepts, image for adversarial ones---validates the dual-probe design and applies symmetrically when \ours{} is run in hybrid mode, since probe selection is a pack property independent of the How update.

\paragraph{Family vs.\ Single Exemplars.}
Table~\ref{table:ablation} compares family-grouped and single-exemplar variants. Family grouping provides large gains on semantically diverse concepts: MJA-Violent (+21.0\%p, from 29.0\% to 50.0\%) and MJA-Illegal (+20.0\%p, from 80.0\% to 100.0\%), where visual sub-categories (\eg, weapons vs.\ physical assault, drug use vs.\ theft) differ substantially. For focused concepts like nudity (88.6\% both) and disturbing (99.0\% both), family and single perform identically.

\input{table/probe_ablation}

% ===========================================================================
\subsection{Cross-Backbone Validation}
\label{subsec:cross_backbone}

A direct consequence of the parameterization-invariance argument in \S\ref{subsec:when} is that \ours{} should transfer to non-$\epsilon$-prediction backbones with no algorithmic change. We test this on SD~3 (MMDiT, velocity parameterization) and FLUX.1-dev (DiT, flow matching with embedded guidance); Table~\ref{table:cross_backbone} reports MJA SR per concept on all three backbones. Only the Where probe is re-implemented per-backbone---cross-attention for SD~v1.4 UNet, joint-attention sub-matrix for SD~3 MMDiT, and a DiT-compatible spatial probe for FLUX. The CAS gate (Eq.~\ref{eq:cas}) and both How updates (Eqs.~\ref{eq:anchor_inpaint}--\ref{eq:hybrid}) are reused \emph{verbatim}.

\input{table/cross_backbone}

\paragraph{What transfers.}
On FLUX.1-dev, the same two-mode trade-off we observed on SD~v1.4 replicates: hybrid dominates on concepts with a visually crisp paired anchor (sexual, disturbing), while anchor inpainting dominates on concepts with ambiguous anchors (violent, illegal). This is strong evidence that the trade-off is a property of the \emph{concept pack and CFG geometry}, not of the $\epsilon$-prediction UNet. On SD~3 (MMDiT), only the anchor row is reported in this revision: the CAS threshold $\tau{=}0.6$ was calibrated for $\epsilon$-prediction and lies above SD~3's velocity-parameterization CAS range ($0.03$--$0.51$), so hybrid guidance is effectively never triggered under that threshold. A per-backbone $\tau$ retune is in progress (we expect $\tau\in[0.2,0.4]$ for SD~3) and the corresponding cells will be filled when stable; we intentionally do \emph{not} report contaminated SD~3 hybrid numbers here.

\paragraph{What this validates.}
The take-away is architectural rather than numerical: CAS is a cosine of \emph{direction differences}, and both How modes are \emph{linear} in the prediction, so the When and How components are identical up to a per-backbone scale; only Where requires per-backbone re-implementation. Cross-backbone transfer of concept-level trends (image-dominant concepts benefit from the image probe across all three backbones; hybrid-vs-anchor winners align between SD~v1.4 and FLUX) therefore confirms that \ours{} is a \emph{framework}, not a SD~v1.4-specific recipe.

% ===========================================================================
\subsection{Image Quality Preservation}
\label{subsec:quality}

Table~\ref{table:main} (right columns) reports FID on COCO. \ours{} achieves FID 6.78, substantially lower than SLD-Max (18.46) and RECE (10.33), confirming that CAS gating effectively prevents quality degradation on benign prompts. We additionally measure VQAScore~\citep{lin2024evaluating} prompt faithfulness: across all datasets, the average VQA drop is $<$0.05 (e.g., Ring-A-Bell: 0.550 $\to$ 0.529, MJA illegal: 0.475 $\to$ 0.474), demonstrating that \ours{} preserves the semantic content of benign regions while selectively suppressing unsafe ones.

% ===========================================================================
\subsection{Evaluator Validation}
\label{subsec:evaluator}

To validate Qwen3-VL as a reliable safety judge, we benchmark it on MJ-Bench~\citep{chen2024mjbench}, a curated dataset of 672 safety image pairs with human preference annotations.

\paragraph{MJ-Bench Results.}
Using the official MJ-Bench protocol (5-criteria multi-image evaluation), Qwen3-VL achieves competitive performance among open-source VLMs on safety preference accuracy. On the safety-specific evaluation (scale [1-10]), Qwen3-VL achieves 49.7\% w/tie and 88.1\% w/o tie accuracy, demonstrating strong discriminative ability when it can distinguish between image pairs.

\paragraph{Cross-Validator Agreement.}
We independently evaluate all generated images with both Qwen3-VL and LLaVA~\citep{liu2023visual}, computing binary safe/unsafe agreement. The two models achieve \textbf{92.3\%} agreement, confirming that our VLM-based evaluation provides consistent and reliable safety assessments.
