\section{Additional Experimental Details}
\label{app:details}

\subsection{Exemplar Prompt Sets}
For each target concept, we use 16 exemplar prompts to compute the concept direction $d_\text{target}$ and CLIP exemplar embeddings. For nudity, representative prompts include: ``a nude person standing'', ``a naked figure in a painting'', ``explicit nudity photography'', etc. Exemplar prompts for other concepts follow a similar pattern, describing typical instances of the target category.

\subsection{Hyperparameter Settings}
Table~\ref{table:hyperparams} lists the default hyperparameters for each concept.

\begin{table}[h]
\centering
\caption{Default hyperparameters per concept.}
\label{table:hyperparams}
\small
\begin{tabular}{lcccc}
\toprule
Concept & CAS $\tau$ & Safety Scale & Target Scale & Anchor Scale \\
\midrule
Nudity & 0.6 & 1.2 & 15.0 & 15.0 \\
Violence & 0.4 & 1.5 & 15.0 & 15.0 \\
Harassment & 0.4 & 1.2 & 15.0 & 15.0 \\
Hate & 0.4 & 1.0 & 15.0 & 15.0 \\
Shocking & 0.6 & 0.8 & 15.0 & 15.0 \\
Illegal & 0.5 & 0.8 & 15.0 & 15.0 \\
Self-harm & 0.4 & 0.8 & 15.0 & 15.0 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{MJ-Bench Evaluation Protocol}
We follow the official MJ-Bench evaluation protocol using the 5-criteria multi-image prompt (prompt\_multi\_image\_number from the MJ-Bench codebase). For each of the 672 safety pairs, both images are presented simultaneously to Qwen3-VL-8B-Instruct, which rates each on Alignment, Aesthetics, Artifacts, Safety, and Bias \& Fairness (1--5 scale), then selects a ``Better Image'' preference. Accuracy is computed by comparing the model's Overall score preference against the ground-truth label, with ties counted as incorrect for w/tie metric and excluded for w/o tie metric.

\subsection{Per-Dataset Detailed Results}
\label{app:detailed}

\paragraph{P4D-N Results.}
On the P4D-N benchmark (151 adversarial prompts), \ours{} achieves 93.4\% SR with the text probe (anchor inpainting, $s_\text{safe}=1.2$) and 92.7\% SR with the dual probe. This compares favorably against SLD-Max (72.0\%) and SAFREE (91.0\%).

\paragraph{I2P Full-Category Breakdown.}
Table~\ref{table:multi} in the main text reports the best-configuration SR for each I2P category. The configurations were selected based on concept-specific validation: concepts with strong textual signals (violence, harassment) benefit from text-only probes, while visually distinctive concepts (nudity, shocking) achieve the best performance with the dual (text + image) probe.

\subsection{MJA Adversarial Benchmark}
\label{app:mja}

The MJA (Metaphor-based Jailbreaking Attack)~\citep{zhang2024metaphor} benchmark uses indirect, metaphorical language to trigger unsafe content generation, making it substantially harder than I2P. Table~\ref{table:mja} compares \ours{} against the SD v1.4 baseline and SAFREE.

\begin{table}[h]
\centering
\caption{MJA adversarial benchmark results (SR\% $\uparrow$, 100 prompts per category).}
\label{table:mja}
\small
\begin{tabular}{lcccc}
\toprule
Concept & Baseline & SAFREE & \textbf{Ours} & $\Delta$ vs SAFREE \\
\midrule
Sexual    & 47.0 & 48.0 & \textbf{74.0} & +26.0 \\
Violent   & 13.0 & 52.0 & \textbf{59.0} & +7.0 \\
Disturbing & 74.0 & \textbf{95.0} & 88.0 & $-$7.0 \\
Illegal   & 60.0 & 77.0 & \textbf{78.0} & +1.0 \\
\bottomrule
\end{tabular}
\end{table}

\ours{} wins on 3 out of 4 MJA categories. The particularly large gain on sexual (+26.0\%p) demonstrates that our dual-probe mechanism detects metaphorical nudity prompts that SAFREE's token projection misses. SAFREE's advantage on disturbing prompts suggests that global suppression can be effective when the entire image, rather than a localized region, contains unsafe content.

\subsection{Concept-Specific Probe Dominance}
\label{app:probe_dominance}

Table~\ref{table:probe_concept} reveals that each probe mode has distinct strengths depending on the concept type.

\begin{table}[h]
\centering
\caption{Probe mode SR\% by concept on MJA (anchor inpainting). Bold = dominant single probe.}
\label{table:probe_concept}
\small
\begin{tabular}{lcccc}
\toprule
Concept & Text & Image & Both & Dominant \\
\midrule
Sexual    & \textbf{65.0} & 56.0 & \textbf{74.0} & Text \\
Violent   & 33.0 & \textbf{45.0} & \textbf{59.0} & Image \\
Disturbing & 79.0 & \textbf{81.0} & \textbf{88.0} & Image \\
Illegal   & 65.0 & \textbf{78.0} & 77.0 & Image \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Nudity/Sexual}: Text probe dominates because even adversarial nudity prompts contain identifiable keywords (``nude'', ``naked'', ``bare'').
\textbf{Violence/Disturbing/Illegal}: Image probe dominates because MJA uses metaphorical descriptions that evade keyword-based detection, but produce visually recognizable patterns that CLIP exemplar embeddings capture.
\textbf{Both always best or tied}: The union fusion ensures robustness against both explicit and implicit adversarial strategies, validating the complementary dual-probe design.

\subsection{Family-Grouped vs Single Exemplar}
\label{app:family}

\begin{table}[h]
\centering
\caption{Family-grouped vs single exemplar SR\% on I2P (both probe, anchor inpainting).}
\label{table:family}
\small
\begin{tabular}{lccc}
\toprule
Concept & Family & Single & $\Delta$ \\
\midrule
Hate      & \textbf{85.3} & 81.4 & +3.9 \\
Illegal   & \textbf{74.6} & 69.4 & +5.2 \\
Harassment & \textbf{87.0} & 86.4 & +0.6 \\
Violence  & 89.2 & 89.8 & $-$0.6 \\
Self-harm & 73.0 & 74.3 & $-$1.3 \\
\bottomrule
\end{tabular}
\end{table}

Family grouping provides meaningful gains for semantically \emph{diverse} concepts (hate, illegal) where sub-categories have distinct visual signatures. For semantically narrow concepts (violence, self-harm), a single exemplar set already captures the concept well.

\subsection{Additional Qualitative Examples}
\label{app:qualitative}

% Placeholder for qualitative comparison figures
% To be added with generated image grids
